An Open-Source Implementation of the CMPS Algorithm for Assessing  Similarity of Bullets

### 2021-06-08 CMPSpaper_writing/CMPSpaper_writing.Rmd
change the usage of level/scale

        -   If `npeaks.set` is an integer vector of length more than 1, for example `npeaks.set = c(5, 3, 1)`, the multi-segment lengths strategy will be used: find the top five peaks \hh{at the first level}, double the segment length and recalculate the cross-correlation function, then identify 3 peaks \hh{at the second level} and repeat the process for a basis segment that is doubled again to identify 1 peak at the third level. \autoref{fig:seg_all} shows an example of 3 levels of basis segment 6 and their corresponding cross-correlation curves and identified peaks.
        -   `get_ccr_peaks(comp, segments, seg_scale, nseg = 1, npeaks = 5)` computes the cross-correlation curve between a basis or scaled segment and the comparison signature and finds peaks in the cross-correlation curve. The number of peaks detected is equal to `npeaks`, which is an integer. `segments`, `seg_scale`, and `nseg` determines the segment in the cross-correlation computation, and `comp` gives the comparison signature. If the multi-segment lengths strategy is used, then `get_ccr_peaks(...)` is called in a `lapply()` for each scale of the basis segment. The resulting list is called `ccr.list`.
    +  **multi-segment lengths strategy:** with the multi-segment lengths strategy being used, a position is identified as a candidate position for registration and is called as a "consistent correlation peak" if it results in a top peak in the cross-correlation curve with a tolerance zone determined by `Tx` in all segment scales. Note that in @cmps, a segment at its largest scale always identifies one peak.
        -   the scaled version of a basis segment is obtained by using function `get_seg_scale(segments, nseg, scale = 2)`. `segments` is a list containing all basis segments, generated by the function `get_segs(...)` in step 1; `nseg` is an integer indicating which basis segment the function should work with. `scale=1` gives the original length of the basis segment. Increasing `scale` by 1 will double the segment length. 

@@

Note that there are \hh{several} parameters in the CMPS algorithm \hh{that are left to} the users \hh{to decide}, such as the length of basis segments `seg_length` in step 1, the number of peaks `npeaks.set` identified on each scale level in step 2, and the length of the tolerance zone `Tx` in both step 2 and 3. 
In our implementation of the CMPS algorithm we used the parameters given in the original CMPS paper [@cmps] as the default values for these parameters. \hh{However, the authors state that no cross-validation has been done - there might be also issues with respect to resolution of the scans. Further research is needed, until then users are advised to think of default values as starting values and consider alternatives.}
<!--Additionally, as shown in @cmps, using multi-segment lengths strategy reduces the number of "false positive" peaks in the cross-correlation curves, thus providing a more conservative way of identifying candidate positions.-->
The main function that combines all steps in the CMPS algorithm described above is called `extract_feature_cmps(...)`. Here we present it with its default parameters.

@@

# algorithm with multi-peak insepction at three different segment scales
cmps_with_multi_scale <- 
  extract_feature_cmps(sigs1$sig, sigs2$sig, 
                       npeaks.set = c(5,3,1), include = "full_result")

@@

In the first example, `npeaks.set` is a vector of three integers, i.e.  the algorithm uses the multi-segment lengths strategy to create the result object `cmps_with_multi_scale`.  For `cmps_without_multi_scale` each basis segment is linked to the top 5 candidate positions. 
We use `include = "full_result"` to capture all results. In this example, the CMPS is 9 when using multiple segments, and 12 when using a single segment. As discussed in @cmps, using multi-segment lengths strategy can reduce the number of false positives when identifying candidate positions; however, any score based method is walking the line between false positives and false negatives. As the number of false positives is reduced the number of false negatives might rise. More discussion and comparisons between the two versions of the CMPS algorithm will be presented in later sections. Note, that the multi scale method is slower because the algorithm is run once for each segment length.

@@

While `cmps_signature_plot()` focuses on the signature level,  `cmps_segment_plot()` focuses on the segment level. It provides the "full result" of `extract_feature_cmps()`, but also takes an argument, `seg.idx`, indicating which segment should be inspected.  When checking `sig.plot$seg_shift` we notice that  segment number 6 is not one of the congruent matching profile segments. We can therefore set `seg.idx = 6` in `cmps_segment_plot()` and investigate the reason why this segment disagrees with the congruent registration position.
For each segment scale, we have two plots: `segment_plot` and `scale_ccf_plot`, as shown in \autoref{fig:segplots} for the example of segment number 6:

*   \autoref{fig:segplot} is the `segment_plot` for basis segment 6 at level one (the top level). We used  `npeaks.set = c(5, 3, 1)` in `extract_feature_cmps()` when calculating the CMPS score. Therefore the top five peaks are identified in the cross-correlation curve at level one. Segment 6 is plotted at the positions where these 5 peaks are identified with dashed lines in the `segment_plot`. The solid black line shows the segment at its original position (which in this example is very close to the actual registration position).

@@

*   \autoref{fig:segplot2} is the `scale_ccf_plot` of basis segment 6 at scale one. It shows the cross-correlation curve computed by the comparison signature and the scale-one basis segment 6. The five highest peaks are marked by dots on the curve.

\begin{figure}
\begin{subfigure}[t]{\textwidth}
\caption{\label{fig:segplot}segment\_plot for segment 6 at scale 1. The original position of segment 6 is indicated by the solid black line. Positions where the segment achieves the 5 highest cross-correlations are indicated by the dashed line segments.}
```{r segplot, opts.label="codefig", echo=TRUE, out.width='\\textwidth', warning = FALSE}
seg.plot <- cmps_segment_plot(
  cmps_with_multi_scale, 
  seg.idx = 6
)
seg.plot[[1]]$segment_plot
```
\end{subfigure}
\begin{subfigure}[t]{\textwidth}
\caption{\label{fig:segplot2}scale\_ccf\_plot shows the cross-correlation curve between the comparison signature and segment 6 at level 1. The five highest peaks are marked by dots. The vertical red dashed line indicates the congruent registration position; the green dashed line shows a peak position in the highest segment level; the blue dashed lines show the tolerance zone around the green dashed line. We can see that none of the five highest peaks at level one falls within the tolerance zone, indicating that there is no consistent correlation peak or a candidate position identified by basis segment 6 under the multi-segment lengths strategy. Thus, the basis segment 6 doesn't vote for the congruent registration position and is not a cmps.}
```{r segplot2, opts.label="codefig", echo=TRUE, out.width='\\textwidth', warning = FALSE}
seg.plot[[1]]$scale_ccf_plot
```
\end{subfigure}
\caption{\label{fig:segplots} Two plots used to investigate basis segment 6 at level one}
\end{figure}

@@

ggarrange(
  plotlist = 
    unlist(seg.plot, 
           recursive = FALSE),
  ncol = 2, 
  nrow = 3)
```
\caption{\label{fig:seg_all} Put segment\_plot and scale\_ccf\_plot of all three scales together. We are identifying five highest peaks at scale one, three peaks at scale two, and one peak at scale three since npeaks.set = c(5, 3, 1). The highest peak position at scale three is marked by the green dashed line across all segment scales, but it doesn't achieve the top five highest peaks at scale one, indicating that there is no consistent correlation peak or a candidate position for basis segment 6 under the multi-segment lengths strategy.}
\end{figure}

In \autoref{fig:seg_all}, the red vertical dashed line indicates the congruent registration position. We can see that the basis segment 6 does obtain a peak near the congruent registration position at scale two and scale three, respectively; however, this position doesn't give one of the five highest peaks at scale one. As a result, segment 6 fails to identify the consistent correlation peak (ccp) and fails to become one of the congruent matching profile segments according to the multi-segment lengths strategy. The identified top five peaks at scale one are also examples of "false positive" peaks. The "true positive" peak (the peak within the tolerance zone of the congruent registration position) is identified at scale two and three by increasing the segment length, which justifies the usage of multi-segment lengths strategy.

@@

Additionally, we applied the CMPS package to two datasets in Hamby study (Hamby set 252 and Hamby set 44), and the results we obtained were promising since we were able to reproduce the results in @cmps qualitatively and achieve a clear separation between the known-match (KM) comparisons and known-nonmatch (KNM) comparisons. However, the results of Hamby set 44 suggested that the CMPS method is sensitive to the choice of parameters in the algorithm, such as the length of basis segments and the number of peaks detected at each segment scale. This indicates that further research is needed to provide a comprehensive cross-validation on parameters used the algorithm. Moreover, we might want to compare the two versions of the CMPS algorithm in the future as well.

>>>>

        -   If `npeaks.set` is an integer vector of length more than 1, for example `npeaks.set = c(5, 3, 1)`, the multi-segment lengths strategy will be used: find the top five peaks at the first level, double the segment length and recalculate the cross-correlation function, then identify 3 peaks at the second level and repeat the process for a basis segment that is doubled again to identify 1 peak at the third level. \autoref{fig:seg_all} shows an example of three levels of basis segment 6 and their corresponding cross-correlation curves and identified peaks.
        -   `get_ccr_peaks(comp, segments, seg_scale, nseg = 1, npeaks = 5)` computes the cross-correlation curve between a basis or scaled segment and the comparison signature and finds peaks in the cross-correlation curve. The number of peaks detected is equal to `npeaks`, which is an integer. `segments`, `seg_scale`, and `nseg` determines the segment in the cross-correlation computation, and `comp` gives the comparison signature. If the multi-segment lengths strategy is used, then `get_ccr_peaks(...)` is called in a `lapply()` for each level of the basis segment. The resulting list is called `ccr.list`.
    +  **multi-segment lengths strategy:** with the multi-segment lengths strategy being used, a position is identified as a candidate position for registration and is called as a "consistent correlation peak" if it results in a top peak in the cross-correlation curve with a tolerance zone determined by `Tx` in all segment levels Note that in @cmps, a segment at its largest scale (highest level) always identifies one peak.
        -   the function `get_seg_scale(segments, nseg, scale = 2)` is used to obtain the scaled version of a basis segment. `segments` is a list containing all basis segments, generated by the function `get_segs(...)` in step 1; `nseg` is an integer indicating which basis segment the function should work with. `scale=1` gives the original length of the basis segment. Increasing `scale` by 1 will double the segment length. 

@@

Note that there are \hh{several} parameters in the CMPS algorithm \hh{that are left to} the users \hh{to decide}, such as the length of basis segments `seg_length` in step 1, the number of peaks `npeaks.set` identified on each level in step 2, and the length of the tolerance zone `Tx` in both step 2 and 3. 
In our implementation of the CMPS algorithm we used the parameters given in the original CMPS paper [@cmps] as the default values for these parameters. \hh{However, the authors state that no cross-validation has been done - there might be also issues with respect to resolution of the scans. Further research is needed, until then users are advised to think of default values as starting values and consider alternatives.}
<!--Additionally, as shown in @cmps, using multi-segment lengths strategy reduces the number of "false positive" peaks in the cross-correlation curves, thus providing a more conservative way of identifying candidate positions.-->
The main function that combines all steps in the CMPS algorithm described above is called `extract_feature_cmps(...)`. Here we present it with its default parameters.

@@

# algorithm with multi-peak insepction at three different segment levels
cmps_with_multi_scale <- 
  extract_feature_cmps(sigs1$sig, sigs2$sig, 
                       npeaks.set = c(5,3,1), include = "full_result")

@@

In the first example, `npeaks.set` is a vector of three integers, i.e.  the algorithm uses the multi-segment lengths strategy to create the result object `cmps_with_multi_scale`.  For `cmps_without_multi_scale` each basis segment is linked to the top 5 candidate positions. 
We use `include = "full_result"` to capture all results. In this example, the CMPS is 9 when using multiple segments, and 12 when using a single segment. As discussed in @cmps, using multi-segment lengths strategy can reduce the number of false positives when identifying candidate positions; however, any score based method is walking the line between false positives and false negatives. As the number of false positives is reduced the number of false negatives might rise. More discussion and comparisons between the two versions of the CMPS algorithm will be presented in later sections. Note, that the multi-segment lengths method is slower because the algorithm is run once for each segment length.

@@

While `cmps_signature_plot()` focuses on the signature level,  `cmps_segment_plot()` focuses on the segment level. It provides the "full result" of `extract_feature_cmps()`, but also takes an argument, `seg.idx`, indicating which segment should be inspected.  When checking `sig.plot$seg_shift` we notice that  segment number 6 is not one of the congruent matching profile segments. We can therefore set `seg.idx = 6` in `cmps_segment_plot()` and investigate the reason why this segment disagrees with the congruent registration position.
\autoref{fig:segplots}  shows the top five registration positions for segment number 6 overlaid on top of the reference signature using hashed lines, the solid line in black shows the original position of the segment (which in this instance is close to the actual registration position). 
Then at each segment level, we will have two plots, `segment_plot` and `scale_ccf_plot`.

*   \autoref{fig:segplot} is the `segment_plot` of basis segment 6 at level one (in its original length). Since when computing the CMPS score we used `npeaks.set = c(5, 3, 1)` in `extract_feature_cmps()`, 5 peaks are identified in the cross-correlation curve at level one, and segment 6 at level one is plotted at the positions where these 5 peaks are identified with dashed lines in the `segment_plot`.

@@

*   \autoref{fig:segplot2} is the `scale_ccf_plot` of basis segment 6 at level one. It shows the cross-correlation curve computed by the comparison signature and the level-one basis segment 6. The five highest peaks are marked by dots on the curve.

\begin{figure}
\begin{subfigure}[t]{\textwidth}
\caption{\label{fig:segplot}segment\_plot for segment 6 at level one. The original position of segment 6 is indicated by the solid black line. Positions where the segment achieves the 5 highest cross-correlations are indicated by the dashed line segments.}
```{r segplot, opts.label="codefig", echo=TRUE, out.width='\\textwidth', warning = FALSE}
seg.plot <- cmps_segment_plot(
  cmps_with_multi_scale, 
  seg.idx = 6
)
seg.plot[[1]]$segment_plot
```
\end{subfigure}
\begin{subfigure}[t]{\textwidth}
\caption{\label{fig:segplot2}scale\_ccf\_plot shows the cross-correlation curve between the comparison signature and segment 6 at level one. The five highest peaks are marked by dots. The vertical red dashed line indicates the congruent registration position; the green dashed line shows a peak position in its highest segment level; the blue dashed lines show the tolerance zone around the green dashed line. We can see here none of the five highest peaks at level one is within the tolerance zone, indicating that there is no consistent correlation peak or a candidate position identified by basis segment 6 under the multi-segment lengths strategy. Thus, the basis segment 6 doesn't vote for the congruent registration position and is not a cmps.}
```{r segplot2, opts.label="codefig", echo=TRUE, out.width='\\textwidth', warning = FALSE}
seg.plot[[1]]$scale_ccf_plot
```
\end{subfigure}
\caption{\label{fig:segplots} Two plots used to investigate basis segment 6 at level one}
\end{figure}

@@

ggarrange(
  plotlist = 
    unlist(seg.plot, 
           recursive = FALSE),
  ncol = 2, 
  nrow = 3)
```
\caption{\label{fig:seg_all} Put segment\_plot and scale\_ccf\_plot of all three levels together. We are identifying five highest peaks at level one, three peaks at level two, and one peak at level three since npeaks.set = c(5, 3, 1). The highest peak position at level three is marked by the green dashed line across all segment levels, but it doesn't achieve the top five highest peaks at level one, indicating that there is no consistent correlation peak or a candidate position for basis segment 6 under the multi-segment lengths strategy.}
\end{figure}

In \autoref{fig:seg_all}, the red vertical dashed line indicates the congruent registration position. We can see that the basis segment 6 does obtain a peak near the congruent registration position at level two and level three, respectively; however, this position doesn't give one of the five highest peaks at level one. As a result, segment 6 fails to identify the consistent correlation peak (ccp) and fails to become one of the congruent matching profile segments according to the multi-segment lengths strategy. The identified top five peaks at level one are also examples of "false positive" peaks. The "true positive" peak (the peak within the tolerance zone of the congruent registration position) is identified at level two and three by increasing the segment length, which justifies the usage of multi-segment lengths strategy.

@@

Additionally, we applied the CMPS package to two datasets in Hamby study (Hamby set 252 and Hamby set 44), and the results we obtained were promising since we were able to reproduce the results in @cmps qualitatively and achieve a clear separation between the known-match (KM) comparisons and known-nonmatch (KNM) comparisons. However, the results of Hamby set 44 suggested that the CMPS method is sensitive to the choice of parameters in the algorithm, such as the length of basis segments and the number of peaks detected at each segment level. This indicates that further research is needed to provide a comprehensive cross-validation on parameters used the algorithm. Moreover, we might want to compare the two versions of the CMPS algorithm in the future as well.
